---
title: "Untitled"
author: "Emerald"
date: "2024-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here)
library(lme4)
library(lmtest)
library(splines)
```



```{r}
df <- read.csv(here("panss1.csv"))
```

## Question

Here, the question would be whether risperidone (for the lowest dose, 2mg) is different from the placebo at reducing negative symptoms of schizophrenia. Is there evidence to say that risperidone (2mg) better than placebo in the mean population negative symptoms over 8 weeks?

## Outcome

The outcome variable in this analysis is the negative subscore. 

```{r}
## choose the data related to our questions
df_clean <- df %>%
  select(-inv,-pospan,-genpan,-totpan,-period) %>%
  filter(treatmnt=="PLACEBO" | treatmnt=="RISPERIDONE_2MG") %>%
  filter(time != -1)
```

## Spaghetti plot
```{r}
df_clean$time_jitter <- df_clean$time + 0.5*(runif(length(df_clean$time)) - 0.5) + 0.2*ifelse(df_clean$treatmnt=="PLACEBO",0,1)
df_clean$trt_col=ifelse(df_clean$treatmnt=="PLACEBO","#EE7E77","#68A7BE")

spagh_plot1 <- ggplot() +
  geom_point(data=df_clean,aes(x=time_jitter,y=negpan,group=treatmnt,col=trt_col),alpha=0.5) +
  geom_line(data=df_clean,aes(x=time_jitter,y=negpan,group=id,col=trt_col),alpha=0.3) +
  ggtitle("Negpan over time")  + xlab("Time") + theme(legend.position='none')
print(spagh_plot1)
ggsave("sp1.png", plot = spagh_plot1, dpi = 500, width = 8, height = 6)

```

## fit random effect model
```{r}
library(lmerTest)
## model1: assuming linear
model1 <- lmer(data=df_clean, negpan~time + time:treatmnt + (time|id), REML = FALSE)
summary(model1)
## model2: with natural spline
model2 <- lmer(data = df_clean, negpan~ns(time,3)+ns(time,3):treatmnt+(time|id), REML = FALSE)
summary(model2)
## null model:
model.null <- lmer(data=df_clean, negpan~ns(time,3) + (time|id), REML = FALSE)
summary(model.null)

## whether there is treatment effect
lrtest(model2, model.null)
## whether natural spline improve the fit
lrtest(model1, model2)

```

## get the predicted negspan
```{r}
df_clean$pred1 = predict(model1, re.form=NA)
df_clean$pred2 = predict(model2, re.form=NA)
df_clean$pred2_re = predict(model2, re.form=NULL)
```

## plot using the fitting model
```{r}
spagh_plot2 <- ggplot() +
  geom_point(data=df_clean,aes(x=time_jitter,y=negpan,group=treatmnt,col=trt_col),alpha=0.2) +
  geom_line(data=df_clean,aes(x=time_jitter,y=negpan,group=id,col=trt_col),alpha=0.1) +
  geom_line(data=df_clean,aes(x=time,y=pred2,group=treatmnt,col=trt_col),alpha=5) + 
  theme(legend.position='none') + ggtitle("Spaghetti plot with LMM fit") + 
  xlab("Time") + theme(legend.position='none')
print(spagh_plot2)
ggsave("sp2.png", plot = spagh_plot2, dpi = 500, width = 8, height = 6)


df_clean$time_shift <- df_clean$time + ifelse(df_clean$treatmnt=="PLACEBO",-0.1,.1)
spagh_plot3=ggplot() +
  geom_point(data=df_clean,aes(x=time_jitter,y=negpan,group=treatmnt,col=trt_col),alpha=0.2) +
  geom_line(data=df_clean,aes(x=time_jitter,y=negpan,group=id,col=trt_col),alpha=0.1) +
  geom_line(data=df_clean,aes(x=time,y=pred2,group=treatmnt,col=trt_col)) + 
  geom_boxplot(data=df_clean,aes(x=time,y=negpan,group=time_shift,col=trt_col)) +
  theme(legend.position='none') + ggtitle("Spaghetti plot with Boxplots and LMM fit") +
  xlab("Time") + theme(legend.position='none')
print(spagh_plot3)
ggsave("sp3.png", plot = spagh_plot3, dpi = 500, width = 8, height = 6)

```



## display of autocorrelation
```{r}
# Reshape the data to wide format
df_clean$resid=df_clean$negpan - df_clean$pred2
df_clean_wide <- df_clean %>% select(id, time, resid) %>%
spread(key = time, value = resid, sep = "")

# Generate the scatterplot matrix
df_clean_wide %>% select(-id) %>% pairs


# Generate the correlation matrix
options(digits = 2) 
# the correlation matix
df_clean_wide %>% select(-id) %>% cor(use="pairwise.complete.obs") 

# Generate the variance matrix
df_clean_wide %>% select(-id) %>% cov(use="pairwise.complete.obs") 

```

## Testing whether the two treatment groups have different mean time curves
```{r}
## The covariance matrix for the coefficients is estimated to be: 
round(matrix(summary(model2)$vcov,ncol=7),5)
## The random effects covariance matrix and residual variance estimates are: 
summary(model2)$varcor

A <- matrix(ns(c(0,1,2,4,6,8), Boundary.knots=c(0,8), knots=c(1,4)), ncol=3)
beta_hat <- matrix(summary(model2)$coefficients[5:7,1],ncol=1)
var_beta_hat <- summary(model2)$vcov[5:7,5:7]
fit <- A %*% beta_hat
var_fit <- A %*% var_beta_hat %*% t(A)
se_fit <- sqrt(diag(var_fit))

result <- data.frame(time=c(0,1,2,4,6,8), fit=fit, se_fit=se_fit, ci_l=fit-2*se_fit, ci_u=fit+2*se_fit)

# plot results
plot_difference <- ggplot(data=result,aes(x=time, y=fit)) + geom_point() + geom_line(col="#77B28C") +
  geom_abline(intercept=0,slope=0,col="red") + 
  geom_line(aes(x=time,y=ci_l),col="#F4D03F") +
  geom_line(aes(x=time,y=ci_u),col="#F4D03F") + 
  ylim(-8,15) + ylab("Mean pospan difference") + 
  xlab("Time in weeks")
ggsave("difference.png", plot = plot_difference, dpi = 500, width = 8, height = 6)

```

### Check the robustness of the test of treatment effect
```{r}
library(clubSandwich)
V_robust <- vcovCR(model2, type="CR1",form = "sandwich")

test_stat_robust <- matrix(beta_hat, nrow=1) %*% 
  solve(V_robust[5:7,5:7]) %*% matrix(beta_hat, ncol=1)

test_stat_model <- matrix(beta_hat,nrow=1) %*% 
  solve(var_beta_hat) %*% matrix(beta_hat, ncol=1)

p_robust <- 1-pchisq(as.numeric(test_stat_robust), df=3)
p_model <- 1-pchisq(as.numeric(test_stat_model), df=3)

robust_result <-list(test_model=test_stat_model, p_model=p_model, test_robust=test_stat_robust,
            p_robust = p_robust)

robust_result
```


### Get prediction for an individual
```{r}
# class demo on predicting an individual's symptoms curve
#
# choose a person at random
#
set.seed(123)
id_random = sample(unique(df_clean$id),1)

# get their data

Y_i = matrix(df_clean$negpan[df_clean$id==id_random],ncol=1)
trt_i = ifelse(df_clean$treatmnt[df_clean$id == id_random]=="PLACEBO",1,0)[1]

# t are the times with observed symptom scores

t_i = sort(matrix(df_clean$time[df_clean$id==id_random],ncol=1))
t_new = matrix(c(0,1,2,4,6,8,9,10), ncol=1)

# t-bases is the n_i (times) by 3 (degrees of freedom) matrix of basis vectors for time

t_bases_i = ns(t_i,Boundary.knots=c(0,8), knots=c(1.5,4.5))
t_bases_new = ns(t_new,Boundary.knots=c(0,8), knots=c(1.5,4.5))

# add the column of 1s for the intercept to get the X matrix for person i

X_i= cbind(rep(1,length(t_i)), t_bases_i,trt_i*t_bases_i)

fixed_i = X_i %*% matrix(summary(model2)$coefficients[,1], ncol=1)

o_i = order(t_i)
miny = min(c(Y_i,fixed_i))*.5
maxy = max(c(Y_i,fixed_i))*1.5

X_new_i = cbind(rep(1,length(t_new)), t_bases_new, trt_i*t_bases_new)
fixed_new_i = X_new_i %*% matrix(summary(model2)$coefficients[,1], ncol=1)  #X times beta

# estimate the covariance matrices required for calculation the predictions and their variance matrix
D = matrix(summary(model2)$varcor[[1]],ncol=2)
sig2 = attr(VarCorr(model2), "sc")^2

Z_i=cbind(rep(1,length(t_i)),t_i)
Z_new=cbind(rep(1,length(t_new)),t_new)
Sigma_11_i = Z_i %*% D %*% t(Z_i) + sig2*diag(rep(1,length(t_i)))
Sigma_12_i = Z_i %*% D %*% t(Z_new)
Sigma_21_i = t(Sigma_12_i)
Sigma_22_i = Z_new %*% D %*% t(Z_new)

mean_Y_new_i = fixed_new_i + Sigma_21_i %*% solve(Sigma_11_i) %*% (Y_i - fixed_i)
var_Y_new_i = Sigma_22_i - Sigma_21_i %*% solve(Sigma_11_i) %*% Sigma_12_i
se_Y_new_i = sqrt(diag(var_Y_new_i))

ci_l_i = mean_Y_new_i - 2*se_Y_new_i
ci_u_i = mean_Y_new_i + 2*se_Y_new_i


# create prediction plot

png(here("prediction_plot.png"), width = 500, height = 350)

plot(t_i[o_i], Y_i[o_i], col="black", xlim=c(0,11), ylim=c(miny, maxy), type="b", 
     ylab = "Negative symptoms", xlab = "Time (weeks)")

title_grp <- df_clean$treatmnt[df_clean$id == id_random][1] %>% 
    str_to_lower() %>% 
    str_replace("_", " ")

title(main = paste0("Predicted and observed negative symptoms for patient ", 
                    id_random, "\n(in ", title_grp, " group)"))

lines(t_new, mean_Y_new_i,col="green")

lines(t_new, fixed_new_i, col="blue", type = "b", pch = 18)

for (j in 1:length(t_new)){
  lines(x= rep(t_new[j],2), y=c(ci_l_i[j],ci_u_i[j]),col="green")
}

dev.off()

```

